ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-09-26 paper/appendix.tex
almost

Figures~\ref{fig:mnist-eccco-lenet} to~\ref{fig:mnist-eccco-jem-ens} show examples of counterfactuals for \textit{MNIST} generated by \textit{ECCCo+} for our different models. Original images are shown on the diagonal and the corresponding counterfactuals are plotted across rows. Figures~\ref{fig:mnist-revise-lenet} to~\ref{fig:mnist-revise-jem-ens} show the same examples but for \textit{REVISE}. Both counterfactual generators have access to the same optimizer. While the results for \textit{REVISE} look fairly poor here, we have observed better results for optizers with higher step sizes.

@@

>>>>

Figures~\ref{fig:mnist-eccco-lenet} to~\ref{fig:mnist-eccco-jem-ens} show examples of counterfactuals for \textit{MNIST} generated by \textit{ECCCo+} for our different models. Original images are shown on the diagonal and the corresponding counterfactuals are plotted across rows. Figures~\ref{fig:mnist-revise-lenet} to~\ref{fig:mnist-revise-jem-ens} show the same examples but for \textit{REVISE}. Both counterfactual generators have access to the same optimizer. While the results for \textit{REVISE} look fairly poor here, we have observed better results for optizers with higher step sizes. Note that the seemingly poor performance by \textit{REVISE} upon visual inspection is not driven by a weak surrogate VAE: Figure~\ref{fig:vae-reg} shows image reconstructions generated by the VAE.

% ECCCo params λ=[0.1,0.5,1.0], reg_strength=0.0

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\linewidth]{../www/vae_rec.png}
  \caption{Randomly drawn \textit{MNIST} images and their reconstructions generated by the VAE used by \textit{REVISE}.}\label{fig:vae-rec}
\end{figure}

### 2023-09-26 paper/body.tex
uh
almost

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (\textit{CNN}). Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset. Full details concerning our parameter choices, training procedures and model performance can be found in the appendix.

@@

Looking at the benchmark results presented in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision} we firstly note that although \textit{REVISE} generally performs best, \textit{ECCCo} and in particular \textit{ECCCo+} often approach SOTA performance. Zooming in on the details we observe that \textit{ECCCo} and its variations do particularly well, whenever the underlying model has been explicitly trained to learn plausible representations of the data. For both tabular datasets in Table~\ref{tab:results-tabular}, \textit{ECCCo} improves plausibility substantially compared to the baseline. This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness (see, for example, Tables~\ref{tab:results-moons} and~\ref{tab:results-german-credit} in the appendix).

@@

This work leverages recent advances in energy-based modelling and conformal prediction in the context of Explainable Artificial Intelligence. We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain. Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals if and only if the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis. This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones. While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid base for future work on faithful counterfactual explanations.

>>>>

For the predictive modelling tasks, we use multi-layer perceptrons (\textit{MLP}), deep ensembles, joint energy models (\textit{JEM}) and convolutional neural networks (LeNet-5 \textit{CNN}). Both joint-energy modelling and ensembling have been associated with improved generative properties and adversarial robustness~\citep{grathwohl2020your,lakshminarayanan2016simple}, so we expect this to be positively correlated with the plausibility of \textit{ECCCo}. To account for stochasticity, we generate multiple counterfactuals for each target class, generator, model and dataset. Full details concerning our parameter choices, training procedures and model performance can be found in the appendix.

@@

Looking at the benchmark results presented in Tables~\ref{tab:results-tabular} and~\ref{tab:results-vision} we firstly note that although \textit{REVISE} generally performs best, \textit{ECCCo} and in particular \textit{ECCCo+} often approach SOTA performance. Upon visual inspection of the generated images we actually find that \textit{ECCCo+} performs much better than \textit{REVISE} (see appendix). Zooming in on the details we observe that \textit{ECCCo} and its variations do particularly well, whenever the underlying model has been explicitly trained to learn plausible representations of the data. For both tabular datasets in Table~\ref{tab:results-tabular}, \textit{ECCCo} improves plausibility substantially compared to the baseline. This broad pattern is mostly consistent for all other datasets, although there are notable exceptions for which \textit{ECCCo} takes the lead on both plausibility and faithfulness (see, for example, Tables~\ref{tab:results-moons} and~\ref{tab:results-german-credit} in the appendix).

@@

This work leverages ideas from energy-based modelling and conformal prediction in the context of counterfactual explanations. We have proposed a new way to generate counterfactuals that are maximally faithful to the black-box model they aim to explain. Our proposed generator, \textit{ECCCo}, produces plausible counterfactuals iff the black-box model itself has learned realistic explanations for the data, which we have demonstrated through rigorous empirical analysis. This should enable researchers and practitioners to use counterfactuals in order to discern trustworthy models from unreliable ones. While the scope of this work limits its generalizability, we believe that \textit{ECCCo} offers a solid base for future work on faithful counterfactual explanations.
